---
title: "Reproducible Research -- Project 1"
author: "Jonathan L. Marrero"
date: "February 7, 2016"
output: html_document
---

### Step 1. Loading and Processing Data 

```{r, echo = TRUE}
library(readr)
df <- read_csv("rep_activity.csv")
```

### Step 2. What is mean total number of steps take per day?

```{r, echo = TRUE}
daily_sum <-tapply(df$steps, df$date, sum, na.rm = TRUE)
hist(daily_sum, breaks = 30, main = "Total Steps Taken Daily", xlab = "Total Steps Per Day")
```
 
#### Mean of the total steps taken each day:

```{r, echo = TRUE}
mean(daily_sum)
```

#### Median of the total steps taken each day:

```{r, echo = TRUE}
median(daily_sum)
```

### Step 3. What is the average daily activity pattern?

##### A plot of the average steps taken at 5-minute intervals for all days:

```{r, echo = TRUE}
plot(tapply(df$steps, df$interval, mean, na.rm = TRUE), type = "l", ylab = "Avg. Steps", xlab = "5 Minute Interval", main = c("Average Steps Taken at 5-Minute Intervals", "(24 Hours)"), ylim = c(0,200))
```

##### The 5-minute interval with the highest average steps across all days:

```{r, echo = TRUE}
which.max(tapply(df$steps, df$interval, mean, na.rm = TRUE))
```

### Step 4. Inputing missing values

##### Number of missing values: 

```{r, echo = TRUE}
sum(is.na(df))
```

#### Creating a new dataset replacing the missing step data with the median for that day.

```{r, echo = TRUE}
df2 <- df ## duplicating the original dataset 
avg <- mean(df2$steps, na.rm = TRUE)
for(i in 1:nrow(df2)){
  if (is.na(df2$steps[i]) == "TRUE"){
    df2$steps[i] <- avg
  }
}

## Calculating the daily totals
daily_sum2 <-tapply(df2$steps, df2$date, sum, na.rm = TRUE)
hist(daily_sum2, breaks = 30, main = "Total Steps Taken Daily", xlab = "Total Steps Per Day")

```

##### Mean of total steps taken per day:

```{r, echo = TRUE}
mean(daily_sum2)
```

##### Median of total steps taken per day:

```{r, echo = TRUE}
median(daily_sum2)
```

#### The impact of replacing missing data: 

```{r, echo = TRUE}
par(mfrow=c(1,2), mar = c(5,4,3.5,2))
boxplot(daily_sum, main = "With Missing Data")
boxplot(daily_sum2, main = "Replaced Missing Data")
```

### Step 5. Are there differences in activity patterns between weekdays and weekends?

```{r, echo = TRUE}
library(lattice)

for(i in 1:nrow(df2)){
  if (weekdays(df2$date[i]) %in% c("Sunday","Saturday")) {
    df2$weekday[i] <- "weekend"
  } else {
    df2$weekday[i] <- "weekday"
  }
}

df2$y <- tapply(df2$steps[df2$weekday == "weekend"], df2$interval[df2$weekday == "weekend"], mean, na.rm = TRUE)

df2$wy <- tapply(df2$steps[df2$weekday == "weekday"], df2$interval[df2$weekday == "weekday"], mean, na.rm = TRUE)

xyplot(y + wy ~ interval | weekday, data = df2, type = "l", layout = c(1,2), ylab = "Steps", xlab = "Interval", main = "Avg Steps Weekend vs Weekday")
```